---
title: "RECOVAR: REusing COrrelations for Validation And Regression"
author: "Ian Hussey"
output:
  html_document:
    code_folding: hide
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r include=FALSE}

# formatting options
# set default chunk options
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

# Dependencies

```{r}

# dependencies
library(lavaan)
library(janitor)
library(tidyverse)
library(knitr)
library(kableExtra)
library(Matrix)

```

# Check correlation matrices for impossible values

## Functions

```{r}

vector_to_matrix <- function(correlation_matrix_as_vector, is_lower_triangle, has_diagonal){
  
  require(lavaan)

  mat <- 
    lavaan::getCov(x = correlation_matrix_as_vector,
                   lower = is_lower_triangle,
                   diagonal = has_diagonal) 
  
  return(mat)
}

check_if_positive_definite <- function(matrix){
  
  pd <- all(eigen(matrix)$values > 0)
  
  return(pd)
}

```

## Small matrix

Is not PD

```{r}

vec0 <- c(1.00,  0.79, -0.10,  
                 1.00,  0.66, 
                        1.00)

mat0 <- vector_to_matrix(vec0, is_lower_triangle = FALSE, has_diagonal = TRUE)

mat0 |>
  knitr::kable() |>
  kableExtra::kable_classic(full_width = FALSE)

check_if_positive_definite(mat0)

```


## Larger matrix

Is PD

```{r}

vec1 <- c(1.00,  0.32,  0.34, -0.09,  0.21,  0.30, 
                 1.00,  0.45, -0.54,  0.54,  0.64,
                        1.00, -0.26,  0.32,  0.34, 
                               1.00, -0.52, -0.49, 
                                      1.00,  0.50,
                                             1.00)

mat1 <- vector_to_matrix(vec1, is_lower_triangle = FALSE, has_diagonal = TRUE)

mat1 |>
  knitr::kable() |>
  kableExtra::kable_classic(full_width = FALSE)

check_if_positive_definite(mat1)

```

Is not PD

```{r}

vec2 <- c(1.00,  0.32,  0.34, -0.09,  0.21, -0.90, # this top right corner correlation was change from .30 to -.90
                 1.00,  0.45, -0.54,  0.54,  0.64,
                        1.00, -0.26,  0.32,  0.34, 
                               1.00, -0.52, -0.49, 
                                      1.00,  0.50,
                                             1.00)

mat2 <- vector_to_matrix(vec2, is_lower_triangle = FALSE, has_diagonal = TRUE)

mat2 |>
  knitr::kable() |>
  kableExtra::kable_classic(full_width = FALSE)

check_if_positive_definite(mat2)

```

# Reuse correlation matrices to fit regressions

## Functions

```{r}

regression_from_correlation_matrix <- function(model, cor, n){
  # dependencies
  require(lavaan)
  require(dplyr)
  require(janitor)

  # fit model
  fit <- 
    sem(model       = model, 
        sample.cov  = cor, 
        sample.nobs = n)
  
  # extract results
  res <- 
    parameterEstimates(fit, 
                       remove.nonfree = TRUE) |>
    # drop irrelevant rows
    filter(op != "~~") |>
    # wrangle variable names and round values
    select(predictor = rhs,
           Beta = est,
           Beta_se = se,
           Beta_ci_lower = ci.lower,
           Beta_ci_upper = ci.upper,
           p = pvalue) |>
    mutate(p = ifelse(p < .0001, "< .0001", janitor::round_half_up(p, 4)))
  
  return(res)
}

```

## Correlation table and N extracted from article

Extracted from Haller et al. (2022) 'To Help or Not to Help? Prosocial Behavior, Its Association With Well-Being, and Predictors of Prosocial Behavior During the Coronavirus Disease Pandemic'

```{r}

n <- 9496

name <- c("Prosocial behavior", "Well-being", "Social Support", "Perceived Stress", "Psychological Flexibility", "Positive Affect")

name <- janitor::make_clean_names(name)

correlation <- 
  lavaan::getCov(x = c(
    1.00,  0.32,  0.34, -0.09,  0.21,  0.30, # Prosocial behavior
           1.00,  0.45, -0.54,  0.54,  0.64, # Well-being
                  1.00, -0.26,  0.32,  0.34, # Social Support
                         1.00, -0.52, -0.49, # Perceived Stress
                                1.00,  0.50, # Psychological Flexibility
                                       1.00  # Positive Affect
  ),
  lower = FALSE,
  diagonal = TRUE, 
  names = name
  )

extracted_results <-
  list(
    n = n,
    name = name,
    correlation = correlation
  )

```

## Original model

Described as non-causal:

$prosocial \sim stress + affect + support + flexiblity$

However, note that this model omits the non-continuous demographics variables as correlations were not/could not be reported. 

```{r}

# fit regression using correlation matrix
fit1 <- regression_from_correlation_matrix(
  model = 'prosocial_behavior ~ perceived_stress + positive_affect + social_support + psychological_flexibility',
  cor = extracted_results$correlation,
  n = extracted_results$n
)

# print results
fit1 |>
  # round variables for printing
  mutate(across(c(where(is.numeric), -p), janitor::round_half_up, digits = 2)) |>
  # print html table
  kable() |>
  kable_classic(full_width = FALSE)

```

- Statistically significant results interpreted as evidence for the role of psychological flexiblity in determining prosocial behavior (after controlling for other variables), although authors note this is not causal evidence.

## Alternative model

Which can nudge authors to consider their implicit causal assumptions:

$flexiblity \sim stress + affect + support + prosocial$

```{r}

# fit regression using correlation matrix
fit2 <- regression_from_correlation_matrix(
  model = 'psychological_flexibility ~ perceived_stress + positive_affect + social_support + prosocial_behavior',
  cor = extracted_results$correlation,
  n = extracted_results$n
)

# print results
fit2 |>
  # round variables for printing
  mutate(across(c(where(is.numeric), -p), janitor::round_half_up, digits = 2)) |>
  # print html table
  kable() |>
  kable_classic(full_width = FALSE)

```

- Results are all still staticially significant, but the model represents very different causal assumptions. In this case, and following the authors logic, prosocial behavior (after controlling for other psycholgoical variables) is shown to drive psychological flexiblity. If the authors are genuinely agnostic with regard to the causal relationships beteen the variables, then shuffle the variables for them and show that associations still hold, and encourage the authors to report these models and results instead. Oftentimes, authors' resistence to this serve to elucidate implicit causal assumptions. This can be useful to the explication of these assumptions. 

