---
title: "2-way between subjects ANOVA from summary statistics"
author: "Ian Hussey"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2-way between subjects ANOVA from summary statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}

# formatting options
# set default chunk options
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

# disable scientific notation
options(scipen = 999) 

```

```{r}

# dependencies
library(dplyr)
library(knitr)
library(kableExtra)

source("../R/anova_from_summary_stats.R")

```

# 2-way between subjects ANOVA from summary statistics 

Summary stats are typically rounded for reporting. Recomputed ANOVA results may therefore be inaccurate. We work around this by recomputing F values for the reported summary stats, and also the min and max F values for summary stats that could be rounded to the reported values. P values and partial eta squared ($\eta_p^2$) effect size and its 90% confidence intervals are then computed from those F values. Note that by default 90% CIs are computed rather than 95% CIs because partial eta2 is always positive (see [Laken's blog on this topic](https://daniellakens.blogspot.com/2014/06/calculating-confidence-intervals-for.html)). 

```{r}

# labels for the experimental groups and interaction
labels <- c("groups 1 (A vs B)", "groups 2 (C vs D)", "groups 1 x groups 2 interaction")

# extracted data - must be formatted as matrices
m <- matrix(c(5.00, 2.69,
              4.83, 5.54),
            ncol = 2)

sd <- matrix(c(2.99, 2.57,
               2.71, 1.84),
             ncol = 2)

n <- matrix(c(40, 20,
              35, 10),
            ncol = 2)

# fit anova from summary stats
res <- anova_from_summary_stats(m      = m,
                                sd     = sd,
                                n      = n,
                                labels = labels)

# print results
res |>
  mutate(p = round_half_up_to_char(p, digits = 5)) |>
  mutate_if(is.numeric, janitor::round_half_up, digits = 3) |>
  knitr::kable() |>
  kableExtra::kable_classic(full_width = FALSE)

```



